Molecular dynamics simulation of synchronization in driven particles

### 2021-02-14 manuscript/manuscript_ajp_2020.tex
moved phase into main body

The competition between the driving force and landscape potential
can produce a variety of hopping patterns in the particle motion. 
%In Fig.~\ref{fig:0}
%reaching a minimum position
%of $y/\lambda = 1$
%over the total period $T = 100$.
%In Fig. ~\ref{fig:0}
%the average displacement during one time period is a single wavelength
%of the substrate $y(t_0+T) - y(t_0) = \lambda$.
%single dc case with figure
%dc_current 0.01
%ac_current 0.05
%ac_frequency 0.01
%fp_y 0.1
%
The relative values of $F^{ac}$, $F^{dc}$ and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
To explore the possible hopping patterns,
we sweep through a range of $F^{dc}$ for fixed $F^{ac}$ and $A_p$
to illustrate
distinct oscillation modes. % as a function of $F^{dc}$.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\langle {v}_{y} \rangle$
over a range of driving forces $F^{dc}$.
%as shown
In Fig.~\ref{fig:1} 
we keep $F^{dc}$ constant
for $10^5$ simulation timesteps
and 
measure the average velocity $\langle v_y \rangle $ 
as a function of $F^{dc}$,
with $F^{ac} = 0.05$, $f=0.01$, and the landscape %force $\vec{F}^{l}$
shown in Fig.~\ref{fig:0}.
%We plot the results in Fig.~\ref{fig:1} for 
The modes appear as 
discrete steps in $\langle v_y \rangle$ %occur because
across a range of $F^{dc}$,

%the average velocity of the particle
%is constant 
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle
due to the landscape confinement.  
In Fig.~\ref{fig:1} at %this occurs at 
low $F^{dc}$ the average velocity $\langle v_y \rangle$ is zero.
Since 
$A_p$ is large compared to the extrema of $F^{D}(t)$,
the particle oscillates back and forth
in a single minima with no net velocity.
At higher $F^{dc}$ the particle velocity 
$\langle v_{y} \rangle$ increases in steps of uniform height,
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.
The step width is non-linear 
and can have a variety of
interesting patterns
such as a devil's staircase related to chaotic dynamics \cite{Bak1986}.
We note the individual steps
are not strictly horizontal 
since particles move continuously
through the landscape, so changes in $F^{dc}$
will create linear changes in 
net particle displacement that are observable
within a single step.

%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.

%This synchronization
%of the particle velocity can be measured
%in terms of the landscape period.
%Phase locked steps occur for average displacements integer
%multiples of the substrate period $n\lambda$.

@@

In Section ~\ref{sec:problems}
we explore our model
with suggested exercises for interested readers.
We include analytic exercises to examine the 
to linear drag equation in Ex.~\ref{ex:reynolds}, 
the equation of motion in Ex.~\ref{ex:n2l},
and 
numerical integration techniques in Ex.~\ref{ex:euler}.
We extend the model with
numerical exercises %extend the model
to include finite temperature effects
in Ex.~\ref{ex:brownian}.
We explore changes in the particle motion
with parameter changes in Ex.~\ref{ex:parameters}.
Methods to visualize the dynamics of the particle motion
are in Ex.~\ref{ex:animation}.

\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{sweep_FDC_vs_vx.pdf}
\caption{Average particle velocity  $\langle v_{y} \rangle$ as a function of $F^{dc}$,
  the constant parameter of $F^D(t)$ defined in Eq.~\ref{eq:drive}.
  The remaining parameters in $F^D(t)$ are as in Fig.~\ref{fig:0}, $F^{ac}=0.05$ and $f = 0.01$.  }
\label{fig:1}
\end{figure}
\end{center}

%In the animation available in Ref.~\cite{supp1} the magenta dot represents the average velocity of the particle $\langle v_{y} \rangle$ at which the particle in Fig. 1(a) is moving.

@@

  In Fig.~\ref{fig:hops_back}
  we fix $f=0.01$, $F^{dc}=0.1$, and and $A_p = 0.1$.
  We show the results for increasing $F^{ac} = 0.2, 0.3, 0.4$.
  %of the system for our original parameters.
  Here $n_b$ increases from $n_b = 0$ in (a) to
  $n_b = 2$ in (b) to $n_b = 4$ in (c).  
  \begin{center}
    \begin{figure}[h!]
      \centering
      \includegraphics[width=\columnwidth]{{parameters_fig6}.pdf}
      \caption{
        Motion of a particle with the same values of $F^{dc}$ and frequency as Fig.~\ref{fig:0} and increasing $F^{ac}$.
        (a) The particle moves forward six minima with no steps backward.
        (b) The particle moves forward eight minima with two steps backward.
        (c) The particle moves forward nine minima with four steps backward.
      }
      \label{fig:hops_back}
    \end{figure}
  \end{center}
  %In Fig.~\ref{fig:hops}

@@

\subsection{Phase plots}
  \label{ex:phase}

  Synchronization 
  can be detected with 
  phase
  plots or Lissajous figures
  that 
  %First reported in 1815 by Nathanial Bowditch
  %and later by Lissajous in 1857  \cite{Lissajous1857}.
  %Lissajous figures
  compare two periodic signals
  into a two dimensional plot.
  In a non-driven system, an appropriate
  phase space is $v_y$ versus $y$.  
  %and easily generated with a computer code 
  %or oscilloscope \cite{Tong1997}.
  When a particle velocity
  synchronizes to a particular spatial location
  the dynamics
  appears as 
  %A mode-locked system is confined to 
  a closed loop in  
  phase space, % showing the relationship
  %between two periodic functions
  with patterns determined by the mode. % frequency ratio,
  %with a frequency generator,
  %where two combined frequencies at perpendicular orientations
  %\footnote{The term phase
  %has two scientific meanings:  
  %oscillation phase $\phi_0$ is the constant of the argument
  %in a periodic function
  %$\sin{(\omega t + \phi_0)}$ and phase space is
  %used to characterize oscillating systems with
  %parametric plots.}.
  In a system driven parallel to the $y-$direction,
  the particle position is monotonically increasing
  in time.
  A synchronized driven particle
  repeats a particular velocity
  at the same $y-$position of the particle
  within the substrate period.
  We define a phase variable 
  \begin{equation}
    %(y mod \lambda) / \lambda
    \phi(t) = 2\pi(y(t)-\bar{v}_y t)/\lambda
  \end{equation}
  that is centered about the average particle velocity $\bar{v}_y$
  and normalized by the substrate period $\lambda$ \cite{Juniper2015}.
  We plot this phase
  versus the phase velocity 
  \begin{equation}
    %(y mod \lambda) / \lambda
    \dot{\phi}(t) = 2\pi(v_y(t)-\bar{v}_y/\lambda.
  \end{equation}
  %Juniper %REF 53

  \begin{center}
    \begin{figure}[h!]
      \centering
      \includegraphics[width=\columnwidth]{phase.pdf}
      \caption{
        Phase plot of a particle with the same values of $F^{dc}$ and frequency as Fig.~\ref{fig:0}.  $F^{ac} = 0.2$ and $A_p = $ (a) $0.0$, (b) $0.05$, (c) $0.1$, and (d) $0.2$.
      }
      \label{fig:phase}
    \end{figure}
  \end{center}

>>>>

The competition between the driving force and landscape potential
can produce a variety of hopping patterns in the particle motion. 
%In Fig.~\ref{fig:0}
%reaching a minimum position
%of $y/\lambda = 1$
%over the total period $T = 100$.
%In Fig. ~\ref{fig:0}
%the average displacement during one time period is a single wavelength
%of the substrate $y(t_0+T) - y(t_0) = \lambda$.
%single dc case with figure
%dc_current 0.01
%ac_current 0.05
%ac_frequency 0.01
%fp_y 0.1
%
The relative values of $F^{ac}$, $F^{dc}$ and $A_p$
control the rate and distance a  particle moves 
forward and backward in the landscape.
To explore the possible hopping patterns,
we sweep through a range of $F^{dc}$ for fixed $F^{ac}$ and $A_p$
to illustrate
distinct oscillation modes. % as a function of $F^{dc}$.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\langle {v}_{y} \rangle$
over a range of driving forces $F^{dc}$.
%as shown

  It is useful to compare
  mode-locked quantities %periodic signals
  in a two dimensional phase plot.
  In a non-driven system, an appropriate
  phase space is particle velocity $v_y$ versus position $y$.  
  %and easily generated with a computer code 
  %or oscilloscope \cite{Tong1997}.
  When $v_y$ %a particle velocity
  is mode-locked to %repeat periodically
  a spatial location
  a closed loop appears in  
  $v_y$ vs $y$ space. % showing the relationship
  %between two periodic functions
  The pattern is determined by the mode,
  or frequency ratio,
  so a $1-1$ mode appears as a circle or oval.
  Nodes appear in the oval 
  for higher modes,
  sometimes forming figure-eights
  or other recognizable patterns.
  A system that is not fully phase locked
  will apppear as an unclosed loop.
  Such quasiperiodic systems are
  not fully synchronized.
  %and the shifts in phase space represent
  %small changes in system behavior.
  With a long simulation time,
  a quasiperiodic phase diagram
  fills the plotting region due to the gradual shift
  in the velocity-position relationship.
  %with a frequency generator,
  %where two combined frequencies at perpendicular orientations
  %\footnote{The term phase
  %has two scientific meanings:  
  %oscillation phase $\phi_0$ is the constant of the argument
  %in a periodic function
  %$\sin{(\omega t + \phi_0)}$ and phase space is
  %used to characterize oscillating systems with
  %parametric plots.}.

  In a system driven parallel to the $y-$direction,
  %the particle velocity and position increase
  %in time. % due to $F^{dc}$.
  %Since a system does not form
  %a closed loop in
  %$v_y$ vs $y$ space,
  we define phase variables 
  to account for the net increase in velocity
  and position.
  The phase position is
  defined as 
  %within the substrate period as 
  %We define a phase variable 
  \begin{equation}
    %(y mod \lambda) / \lambda
    \phi(t) = 2\pi(y(t)-\bar{v}_y t)/\lambda
  \end{equation}
  centered about the average particle velocity $\bar{v}_y$
  and normalized by the substrate period $\lambda$ \cite{Juniper2015}.
  We plot this phase
  versus the phase velocity 
  \begin{equation}
    %(y mod \lambda) / \lambda
    \dot{\phi}(t) = 2\pi(v_y(t)-\bar{v}_y/\lambda.
  \end{equation}
  %Juniper %REF 53

  In Fig.~\ref{fig:phase}
  we plot $\dot{\phi}(t)$ versus $\phi(t)$
  for increasing substrate amplitude $A_p$.
  We fix the driving force with parameters
  $F^{dc} = 0.1$ and frequency $f = 0.01$ and $F^{ac} = 0.2$.
  In Fig.~\ref{fig:phase}(a)
  there is no substrate ($A_p = 0.0$)
  and the phase plot is a closed oval
  centered at
  $(\phi,\dot{\phi})=(10,0)$
  with a semimajor axis of $\Delta \phi = 20$ %ANALYTIC
  and semiminor axis of $\Delta \dot{\phi} = 2\pi/\lambda v_{max}$.
  [insert analysis.]
  As the force from the substrate
  increases,
  nodes form in the phase diagram
  and the motion becomes quasiperiodic
  so the loops in phase space do not fully overlap.
  In Fig.~\ref{fig:phase}(b)
  $A_p = 0.05$
  the spatial properties are similar to
  those in $A_p = 0.0$
  with 
  $(\phi,\dot{\phi})=(10,0)$
  with a semimajor axis of $\Delta \phi = 20$.
  The dynamical properties
  are asymmetric,
  with nodes in the positive $\dot{\phi}$ values.
  [insert analysis.]
  As the substrate
  force becomes larger,
  $\phi$ shifts to lower values
  and $\dot{\phi}$ to positive values,
  indicating
  the particle can only move 
  backward over a short distance.
  \begin{center}
    \begin{figure}[h!]
      \centering
      \includegraphics[width=\columnwidth]{phase.pdf}
      \caption{
        Phase plot of a particle with the same values of $F^{dc}$ and frequency as Fig.~\ref{fig:0}.  $F^{ac} = 0.2$ and $A_p = $ (a) $0.0$, (b) $0.05$, (c) $0.1$, (d) $0.2$, (e) $0.25$ and (f) $0.3$.
      }
      \label{fig:phase}
    \end{figure}
  \end{center}

In Fig.~\ref{fig:1} 
we keep $F^{dc}$ constant
for $10^5$ simulation timesteps
and 
measure the average velocity $\langle v_y \rangle $ 
as a function of $F^{dc}$,
with $F^{ac} = 0.05$, $f=0.01$, and the landscape %force $\vec{F}^{l}$
shown in Fig.~\ref{fig:0}.
%We plot the results in Fig.~\ref{fig:1} for 
The modes appear as 
discrete steps in $\langle v_y \rangle$ %occur because
across a range of $F^{dc}$,

%the average velocity of the particle
%is constant 
%despite the overall magnitude
%of $F^{dc}$.
Each step represents a different pattern of hops
between substrate minima
performed by the particle
due to the landscape confinement.  
In Fig.~\ref{fig:1} at %this occurs at 
low $F^{dc}$ the average velocity $\langle v_y \rangle$ is zero.
Since 
$A_p$ is large compared to the extrema of $F^{D}(t)$,
the particle oscillates back and forth
in a single minima with no net velocity.
At higher $F^{dc}$ the particle velocity 
$\langle v_{y} \rangle$ increases in steps of uniform height,
$\langle v_{y} \rangle = n \lambda f$,
where $n$ is an integer.
The step width is non-linear 
and can have a variety of
interesting patterns
such as a devil's staircase related to chaotic dynamics \cite{Bak1986}.
We note the individual steps
are not strictly horizontal 
since particles move continuously
through the landscape, so changes in $F^{dc}$
will create linear changes in 
net particle displacement that are observable
within a single step.
\begin{center}
\begin{figure}[h!]
\centering
\includegraphics[width=\columnwidth]{sweep_FDC_vs_vx.pdf}
\caption{Average particle velocity  $\langle v_{y} \rangle$ as a function of $F^{dc}$,
  the constant parameter of $F^D(t)$ defined in Eq.~\ref{eq:drive}.
  The remaining parameters in $F^D(t)$ are as in Fig.~\ref{fig:0}, $F^{ac}=0.05$ and $f = 0.01$.  }
\label{fig:1}
\end{figure}
\end{center}
%$\lambda = L/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
%of the landscape,
%and $f = 0.01$ cycles per time unit.
%This synchronization
%of the particle velocity can be measured
%in terms of the landscape period.
%Phase locked steps occur for average displacements integer
%multiples of the substrate period $n\lambda$.
%\subsection{Phase plots}
%  \label{ex:phase}

@@

In Section ~\ref{sec:problems}
we explore our model
with suggested exercises for interested readers.
We include analytic exercises to examine the 
to linear drag equation in Ex.~\ref{ex:reynolds}, 
the equation of motion in Ex.~\ref{ex:n2l},
and 
numerical integration techniques in Ex.~\ref{ex:euler}.
We extend the model with
numerical exercises %extend the model
to include finite temperature effects
in Ex.~\ref{ex:brownian}.
We explore changes in the particle motion
with parameter changes in Ex.~\ref{ex:parameters}.
Methods to visualize the dynamics of the particle motion
are in Ex.~\ref{ex:animation}.
%In the animation available in Ref.~\cite{supp1} the magenta dot represents the average velocity of the particle $\langle v_{y} \rangle$ at which the particle in Fig. 1(a) is moving.

@@

  In Fig.~\ref{fig:hops_back}
  we fix $f=0.01$, $F^{dc}=0.1$, and and $A_p = 0.1$.
  We show the results for increasing $F^{ac} = 0.2, 0.3, 0.4$.
  %of the system for our original parameters.
  Here the number of backwards steps
  $n_b$ increases from $n_b = 0$ in (a) to
  $n_b = 2$ in (b) to $n_b = 4$ in (c).  
  \begin{center}
    \begin{figure}[h!]
      \centering
      \includegraphics[width=\columnwidth]{{parameters_fig6}.pdf}
      \caption{
        Motion of a particle with the same values of $F^{dc}$ and frequency as Fig.~\ref{fig:0} and increasing $F^{ac}$.
        (a) The particle moves forward six minima with no steps backward.
        (b) The particle moves forward eight minima with two steps backward.
        (c) The particle moves forward nine minima with four steps backward.
      }
      \label{fig:hops_back}
    \end{figure}
  \end{center}
  %In Fig.~\ref{fig:hops}

@@
